A spiral galaxy's mass distribution uncovered through lensing and dynamics

### 2015-11-13 docs/abstract_v3.01.tex
Corrections in Section 2 and 3

We analyse the stellar and dark matter distribution in the spiral galaxy SDSS J1331+3638 (J1331) by means of two independent methods: gravitational lensing and dynamical modelling. Hubble Space Telescope (HST) imaging by the SWELLS survey shows, that J1331's bulge is superimposed by a quadruplet of extended lensing images. By fitting a gravitational potential model to the image positions, we tightly constrain the mass inside the Einstein radius $R_\text{ein}=(0.91\pm0.02)''(\simeq1.83\pm0.04~\text{kpc})$ to within 4\%: $M_\text{ein} = (7.8\pm0.3) \cdot 10^{10} M_\odot$. According to long-slit major axis stellar kinematics, J1331 has a counter-rotating stellar core inside $\sim 2''(\simeq4.02~\text{kpc})$. We model the observed stellar kinematics in J1331's central regions by finding Multi-Gaussian Expansion (MGE) models for the stellar and dark matter distribution that solve the axisymmetric Jeans equations. We find that J1331 requires a steep total mass-to-light ratio gradient in the center to reproduce the observed stellar kinematics. The best fit dynamical model predicts a total mass inside the Einstein radius consistent with the lens model, and vice versa the lens model gives an successful prediction for the observed kinematics in the galaxy center. For a dynamical model including a NFW dark matter halo,  we constrain the halo to have virial velocity $v_{200} \simeq 240 \pm 40$ km/s and a concentration of $c_{200} \simeq 8 \pm 2$ in case of a moderate tangential velocity anisotropy of $\beta_z \simeq -0.4 \pm 0.1$. The NFW halo models can successfully reproduce the signatures of J1331's counter-rotating stellar core and predict J1331's stellar and gas rotation curve at larger radii. However, all these models are more massive than expected from the stellar velocity dispersion at larger radii alone. We speculate that this could indicate a non-trivial re-distribution of matter due to a possible merger event in J1331's past. This could have resulted in a stellar core in J1331's very center with a Milky-Way-like Kroupa/Chabrier initial mass function (IMF), surrounded by a extended stellar bulge population with an IMF even more bottom-heavy than a Salpeter IMF \Wilma{[TO DO: Word limit is 250.]}.

>>>>

We analyse the stellar and dark matter distribution in the spiral galaxy SDSS J1331+3628 (J1331) by means of two independent methods: gravitational lensing and dynamical modelling. Hubble Space Telescope (HST) imaging by the SWELLS survey shows, that J1331's bulge is superimposed by a quadruplet of extended lensing images. By fitting a gravitational potential model to the image positions, we tightly constrain the mass inside the Einstein radius $R_\text{ein}=(0.91\pm0.02)''(\simeq1.83\pm0.04~\text{kpc})$ to within 4\%: $M_\text{ein} = (7.8\pm0.3) \cdot 10^{10} M_\odot$. According to long-slit major axis stellar kinematics, J1331 has a counter-rotating stellar core inside $\sim 2''(\simeq4.02~\text{kpc})$. We model the observed stellar kinematics in J1331's central regions by finding Multi-Gaussian Expansion (MGE) models for the stellar and dark matter distribution that solve the axisymmetric Jeans equations. We find that J1331 requires a steep total mass-to-light ratio gradient in the center to reproduce the observed stellar kinematics. The best fit dynamical model predicts a total mass inside the Einstein radius consistent with the lens model, and vice versa the lens model gives an successful prediction for the observed kinematics in the galaxy center. For a dynamical model including a NFW dark matter halo,  we constrain the halo to have virial velocity $v_{200} \simeq 240 \pm 40$ km/s and a concentration of $c_{200} \simeq 8 \pm 2$ in case of a moderate tangential velocity anisotropy of $\beta_z \simeq -0.4 \pm 0.1$. The NFW halo models can successfully reproduce the signatures of J1331's counter-rotating stellar core and predict J1331's stellar and gas rotation curve at larger radii. However, all these models are more massive than expected from the stellar velocity dispersion at larger radii alone. We speculate that this could indicate a non-trivial re-distribution of matter due to a possible merger event in J1331's past. This could have resulted in a stellar core in J1331's very center with a Milky-Way-like Kroupa/Chabrier initial mass function (IMF), surrounded by a extended stellar bulge population with an IMF even more bottom-heavy than a Salpeter IMF \Wilma{[TO DO: Word limit is 250.]}.

### 2015-11-13 docs/data_v3.01.tex
Corrections in Section 2 and 3

For our lensing and dynamics analysis of J1331 we use on the one hand Hubble Space Telescope (HST) imaging by \cite{SWELLSI} in two filters (F450W and F814W), and on the other hand line-of-sight stellar kinematics along J1331's major axis by \citet{SWELLSV}.

\paragraph{Redshift.} \Wilma{[TO DO: WHO???]} found from SDSS spectra that J1331 has two redshifts inside $1''$: The smaller one, $z_d = 0.113$, is the redshift of J1331 itself, the larger one, $z_s = 0.254$, is the redshift of the lensed background source \citep{SWELLSIII}. According to its redshift and the WMAP5 \Wilma{[TO DO: Do I have to explain the Acronym???]} cosmology \citep{WMAP5cosm}, J1331 has an angular diameter distance of 414 Mpc, which translates into a transverse scaling of $1'' \hat{=} 2.01~\text{kpc}$.

\paragraph{HST imaging.} We use the follow-up HST imaging, with which \citet{SWELLSI} confirmed that J1331 is indeed a strong gravitational lens \Wilma{[TO DO: Was it actually SWELLS I that confirmed that???]}. They performed high resolution imaging with the Hubble Space Telescope's (HST) Wide-Field Planetary Camera 2 (WFPC2) and its WF3 CCD chip.  The images we are using are a combination of four exposures with each an exposure of 400s and were drizzled to a pixel scale of 1 pixel = $0.05''$. In particular we use the images in the filters F450W, to identify the positions of the bluish lensing images, and F814W to create a surface brightness MGE model of the reddish bulge (I-band), used in the JAM dynamical modelling.

\paragraph{Stellar kinematics.} For the dynamical modelling we also use the stellar kinematics for J1331 measured by \citet{SWELLSV}. They obtained long-slit spectra along J1331's major-axis with the Low Resolution Imaging Spectrograph (LRIS) on the Keck I 10m telescope. The width of the slot was $1''$ and the seeing conditions had a FWHM of $\sim 1.1''$. Spectra for spatial bins of different widths along the major axis were extracted. Analogously to \citet{SWELLSII} they measured line-of-sight rotation velocities ($v_\text{rot}$) and stellar velocity dispersion ($\sigma$) by fitting Gaussian line profiles to emission lines in these spectra. Gas kinematics were extracted from fits to H$\alpha$ and NII lines, as tracers for inonized gas.
\\The stellar kinematics, $v_\text{rot}$, $\sigma$ and $v_\text{rms}^2=v_\text{rot}^2 + \sigma^2$ are shown in Figure \ref{fig:kinematics}. The rotation curve reveals a counter-rotating core within $2''\simeq$ 4 kpc. Outside of $\sim 3.5''$ there is a steep drop in the dispersion, which is exptected at the boundary between the pressure supported bulge and the rotationally supported disk, which appears around this radius in the F450W filter in Figure \ref{fig:F450W}. However, in the brighter F814W filter in Figure \ref{fig:F814W}  the large reddish bulge extends out to $\sim5''$. 
\\Inside of $\sim 4''$, the data appears to be symmetric, outside of this the assumption of axisymmetry seems not to be valid anymore, considering the data. We add -2.3 km/s to the $v_\text{rot}$ to ensure $v_\text{rot}(R'=0) \sim 0$ as a possible correction term for a systematic misjudgement of the systemic velocity. We also symmetrize the data within $4''$ and asign a minimum error of $\delta v_\text{rms} > 5$ km/s to the $v_\text{rms}$ data. In the JAM modelling, which is based on the assumption of axisymmetry, only kinematics within $\sim 2.5$ and $4''$ are used. Another reason to restrict to modelling on the bulge region is that our MGE in Table \ref{tab:MGEF814W} is only a good representation of J1331's F814W light distribution inside $\sim 5''$.\\

\Wilma{[TO DO: How to deal with the typo in J1331s name???? SWELLS call it J1331+3638, but its actual coordinates are 1331+3628.]}

>>>>

For our lensing and dynamics analysis of J1331 we use both Hubble Space Telescope (HST) imaging by \cite{SWELLSI} in two filters (F450W and F814W), as well as line-of-sight stellar kinematics along J1331's major axis by \citet{SWELLSV}.

\paragraph{Redshift.} \citet{SWELLSI} found from SDSS spectra that J1331 has two redshifts inside $1''$: The smaller one, $z_d = 0.113$, is the redshift of J1331 itself, the larger one, $z_s = 0.254$, is the redshift of the lensed background source \citep{SWELLSIII}. According to its redshift and the WMAP5 cosmology \citep{WMAP5cosm}, J1331 has an angular diameter distance of 414 Mpc, which translates into a transverse scaling of $1'' \hat{=} 2.01~\text{kpc}$.

\paragraph{HST imaging.} We use the follow-up HST imaging, with which \citet{SWELLSI} confirmed that J1331 is indeed a strong gravitational lens. They performed high resolution imaging with the Hubble Space Telescope's (HST) Wide-Field Planetary Camera 2 (WFPC2) and its WF3 CCD chip.  The images we are using are a combination of four exposures with each an exposure of 400s and were drizzled to a pixel scale of 1 pixel = $0.05''$. In particular we use the images in the filters F450W, to identify the positions of the bluish lensing images, and F814W to create a surface brightness MGE model of the reddish bulge (I-band), used in the JAM dynamical modelling.

\paragraph{Stellar kinematics.} For the dynamical modelling we also use the stellar kinematics for J1331 measured by \citet{SWELLSV}. They obtained long-slit spectra along J1331's major-axis with the Low Resolution Imaging Spectrograph (LRIS) on the Keck I 10m telescope. The width of the slot was $1''$ and the seeing conditions had a FWHM of $\sim 1.1''$. Spectra for spatial bins of different widths along the major axis were extracted. They measured line-of-sight stellar rotation velocities ($v_\text{rot}$) and stellar velocity dispersion ($\sigma$) by fitting Gaussian line profiles to emission \Wilma{[TO DO: Check if really emission lines or, more likely, absorption lines]} lines in these spectra \Wilma{[TO DO: Find out which lines]}. Gas kinematics were extracted from fits to H$\alpha$ and NII lines, as tracers for ionized gas.
\\The stellar kinematics, $v_\text{rot}$, $\sigma$ and $v_\text{rms}^2=v_\text{rot}^2 + \sigma^2$ are shown in Figure \ref{fig:kinematics}. The rotation curve reveals a counter-rotating core within $2''\simeq$ 4 kpc. Outside of $\sim 3.5''$ there is a steep drop in the dispersion, which is expected at the boundary between the pressure supported bulge and the rotationally supported disk, which appears around this radius in the F450W filter in Figure \ref{fig:F450W}. However, in the brighter F814W filter in Figure \ref{fig:F814W}  the large reddish bulge extends out to $\sim5''$. 
\\Inside of $\sim 4''$, the data appears to be symmetric, outside of this the assumption of axisymmetry seems not to be valid anymore, considering the data. We add -2.3 km/s to the $v_\text{rot}$ to ensure $v_\text{rot}(R'=0) \sim 0$ as a possible correction term for a misjudgement of the systemic velocity. We also symmetrize the data within $4''$ and asign a minimum error of $\delta v_\text{rms} > 5$ km/s to the $v_\text{rms}$ data. In the JAM modelling, which is based on the assumption of axisymmetry, only stellar kinematics with $R'  \lesssim 2.5''$ or $R' \lesssim 4''$ are used. Another reason to restrict to modelling on the bulge region is that our MGE in Table \ref{tab:MGEF814W} is only a good representation of J1331's F814W light distribution inside $\sim 5''$.\\

### 2015-11-13 docs/introduction_v3.01.tex
Corrections in Section 2 and 3

%Dark Matter
The flat rotation curves of galaxies \citep{1978ApJ...225L.107R} were the first indication that galaxies might reside in massive, almost spherical halos made of invisible ``dark'' matter. The standard model of cosmology, based on results by the Planck Mission \citep{WMAP5cosm} \Wilma{[TO DO: Aaron thinks Dunkley is a WMAP paper.... Check difference WMAP/Planck...]}, infers that $\sim 85\%$ of the universe's total matter content is in the form of non-baryonic cold dark matter. Simulations suggest that this kind of dark matter forms cuspy halos following a Navarro-Frenck-White (NFW) profile \citep{NFW96}. However, central dark matter density cusps are not observed in dark matter dominated dwarf galaxies \Wilma{[TO DO: REF: Aaron suggests Moore 1994, de Blok et al. 2001]}. Their existence in more massive galaxies depends strongly on stellar mass-to-light ratio \Wilma{[TO DO: REF: Aaron suggests Dutton et al. 2011 "dark halo response and the IMF"]}. Overall, observations suggest cored dark matter halos. This discrepancy, known as the core-cusp problem, might be resolved by galaxy formation processes such as mergers and outflows (e.g., \citealt{2001ApJ...560..636E} \Wilma{[TO DO: more REF: Aaron suggests Pontzen \& Governato 2012]}).

@@

%Lensing
Strong gravitational lensing is an independent method to tightly constrain the projected total mass of galaxies inside $\sim 1''$ \Wilma{[TO DO: REF: Aaron suggests Treu 2010]}. Massive galaxies can act as gravitational lenses and deflect the light of background sources, giving rise to multiple distorted images of the source \Wilma{[TO DO: REF]}. By 2010 over 200 strong gravitational galaxy lenses had been discovered \citep{2010ARA&A..48...87T} and the number is still rising.

@@

%Characteristics of J1331
One of the SWELLS galaxies is SDSS J1331+3638, to which we refer as J1331 in the remainder of this work. It is a spiral galaxy of approximate Hubble type Sb at right ascension = 202.91800$^\circ$ and a declination = 36.46999$^\circ$ (epoch J2000). It has bluish spiral arms and a large reddish bulge (see Figures \ref{fig:F450W} and \ref{fig:F814W}), which is superimposed by quadruplet of extended bluish images approximately at a distance of $1''$ from the galaxy center (see Figure \ref{fig:lens_just_imgpos}). The lensed object might be a star-forming blob of a background galaxy. J1331 stands out of the SWELLS sample because of its large counter-rotating core (see Figure \ref{fig:kinematics}), which might be an indication that J1331 underwent a minor merger in its recent past.

@@

\begin{figure*}
\centering
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{fig/first_glimpse_450.ps}
  \caption{J1331 in F450W}
  \label{fig:F450W}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{fig/first_glimpse_814.ps}
  \caption{J1331 in F814W}
  \label{fig:F814W}
\end{subfigure}
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{fig/lens_imgpos.ps}
  \caption{The lensing images}
  \label{fig:lens_just_imgpos}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{fig/stellar_kinematics_data.ps}
  \caption{Stellar Kinematics by \citet{SWELLSV}}
  \label{fig:kinematics}
\end{subfigure}
\caption{Hubble Space telescope (HST) images and stellar kinematics of the galaxy SDSS J1331+3638 (J1331), which has a large counter-rotating core and whose bulge acts as a strong lens for a bluish background source. \emph{Panel (a) and (b):} HST/WFPC2/WFC3 images of J1331 by \citet{SWELLSI} in two filters, F450W in panel (a) and F814W in panel (b). The galaxy's coordinates on the sky are right ascension $\alpha$ = 202.91800$^\circ$ and declination $\delta$ = 36.46999$^\circ$ (epoch J2000). Image orientation and scaling are indicated in panel (a); the scaling transformation from arcseconds to the physical size of the galaxy in kpc uses the galaxy's redshift $z_d = 0.113$ \citep{SWELLSIII} (i.e. assumes an angular diameter distance of 414 Mpc). The color scaling of these two images is the same. The black solid line in panel (b) shows the orientation of the major-axis. The line has a length of $10''$ and indicates the region within which we carry out the Jeans modelling. \emph{Panel (c):} The central region of J1331 in F450W, surface brightness subtracted. An IRAF ellipse \Wilma{[TO DO: How to write and reference????]} fit to the F450W surface brightness in panel (a) was subtracted from the image. The (smoothed) residuals within the white square in panel (a) are shown in panel (c). Four bright blobs (A,B,C and D) become visible, which are arranged in a typical strong lensing configuration around the center of the galaxy (G). \emph{Panel (d):} Stellar Kinematics along the galaxy's major axis as measured by \citet{SWELLSV}, line-of-sight rotation velocity $v_\text{rot}$, line-of-sight velocity dispersion $\sigma$ and the rms-velocity $v_\text{rms} = \sqrt{v_\text{rot}^2 + \sigma^2}$. The dotted line in panel (b) indicates the galaxy's effective half-light radius (in the F814W filter), $R_\text{eff} = 2.6" = 5.2$ kpc. The $v_\text{rot}$ curve reveals that J1331 is counter-rotating within $R_\text{eff}$. \Wilma{[TO DO: Replace arcsec with '' in panel d)]} \Wilma{[TO DO: Mention, that the configuration of the additional two bright blobs do not suggest that they form a lensing doublet. They are probably background star formation regions as well. The blob above A could be stretched due to lensing and have a faint counter image between B and G. Maybe indicate in plot?]}}
\label{fig:specialJ1331}
\end{figure*}

>>>>

%Dark Matter
The flat rotation curves of galaxies \citep{1978ApJ...225L.107R} were the first indication that galaxies might reside in massive, almost spherical halos made of invisible ``dark'' matter. The standard model of cosmology, based on results by the Planck Mission \citep{WMAP5cosm} \Wilma{[TO DO: Aaron thinks Dunkley is a WMAP paper.... Check difference WMAP/Planck...]}, infers that $\sim 85\%$ of the universe's total matter content is in the form of non-baryonic cold dark matter. Simulations suggest that this kind of dark matter forms cuspy halos following a Navarro-Frenk-White (NFW) profile \citep{NFW96}. However, central dark matter density cusps are not observed in dark matter dominated dwarf galaxies (e.g., \citealt{1994Natur.370..629M,2001ApJ...552L..23D}). Their existence in more massive galaxies depends strongly on stellar mass-to-light ratio (e.g., \citealt{2011MNRAS.416..322D}). Overall, observations suggest cored dark matter halos. This discrepancy, known as the core-cusp problem, might be resolved by galaxy formation processes such as mergers and outflows (e.g., \citealt{2001ApJ...560..636E,2012MNRAS.421.3464P}).

@@

%Lensing
Strong gravitational lensing is an independent method to tightly constrain the projected total mass of galaxies inside $\sim 1''$. Massive galaxies can act as gravitational lenses and deflect the light of background sources, giving rise to multiple distorted images of the source. By 2010 over 200 strong gravitational galaxy lenses had been discovered and the number is still rising (e.g., \citealt{2010ARA&A..48...87T}).

@@

%Characteristics of J1331
One of the SWELLS galaxies is SDSS J1331+3628, to which we refer as J1331 in the remainder of this work. It is a spiral galaxy of approximate Hubble type Sb at right ascension = 202.91800$^\circ$ and a declination = 36.46999$^\circ$ (epoch J2000). It has bluish spiral arms and a large reddish bulge (see Figures \ref{fig:F450W} and \ref{fig:F814W}), which is superimposed by quadruplet of extended bluish images approximately at a distance of $1''$ from the galaxy center (see Figure \ref{fig:lens_just_imgpos}). The lensed object might be a star-forming blob of a background galaxy. J1331 stands out of the SWELLS sample because of its large counter-rotating core (see Figure \ref{fig:kinematics}), which might be an indication that J1331 underwent a minor merger in its recent past.

@@

\begin{figure*}
\centering
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{fig/first_glimpse_450.ps}
  \caption{J1331 in F450W}
  \label{fig:F450W}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{fig/first_glimpse_814.ps}
  \caption{J1331 in F814W}
  \label{fig:F814W}
\end{subfigure}
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{fig/lens_imgpos.ps}
  \caption{The lensing images}
  \label{fig:lens_just_imgpos}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{fig/stellar_kinematics_data.ps}
  \caption{Stellar Kinematics by \citet{SWELLSV}}
  \label{fig:kinematics}
\end{subfigure}
\caption{Hubble Space telescope (HST) images and stellar kinematics of the galaxy SDSS J1331+3628 (J1331), which has a large counter-rotating core and whose bulge acts as a strong lens for a bluish background source. \emph{Panel (a) and (b):} HST/WFPC2/WFC3 images of J1331 by \citet{SWELLSI} in two filters, F450W in panel (a) and F814W in panel (b). The galaxy's coordinates on the sky are right ascension $\alpha$ = 202.91800$^\circ$ and declination $\delta$ = 36.46999$^\circ$ (epoch J2000). Image orientation and scaling are indicated in panel (a); the scaling transformation from arcseconds to the physical size of the galaxy in kpc uses the galaxy's redshift $z_d = 0.113$ \citep{SWELLSIII} (i.e. assumes an angular diameter distance of 414 Mpc). The color scaling of these two images is the same. The black solid line in panel (b) shows the orientation of the major-axis. The line has a length of $10''$ and indicates the region within which we carry out the Jeans modelling. \emph{Panel (c):} The central region of J1331 in F450W, surface brightness subtracted. An IRAF \emph{ellipse} fit to the F450W surface brightness in panel (a) was subtracted from the image. The (smoothed) residuals within the white square in panel (a) are shown in panel (c). Four bright blobs (A,B,C and D) become visible, which are arranged in a typical strong lensing configuration around the center of the galaxy (G). \emph{Panel (d):} Stellar Kinematics along the galaxy's major axis as measured by \citet{SWELLSV}, line-of-sight rotation velocity $v_\text{rot}$, line-of-sight velocity dispersion $\sigma$ and the rms-velocity $v_\text{rms} = \sqrt{v_\text{rot}^2 + \sigma^2}$. The dotted line in panel (b) indicates the galaxy's effective half-light radius (in the F814W filter), $R_\text{eff} = 2.6" = 5.2$ kpc. The $v_\text{rot}$ curve reveals that J1331 is counter-rotating within $R_\text{eff}$. \Wilma{[TO DO: Replace arcsec with '' in panel d)]} \Wilma{[TO DO: Mention, that the configuration of the additional two bright blobs do not suggest that they form a lensing doublet. They are probably background star formation regions as well. The blob above A could be stretched due to lensing and have a faint counter image between B and G. Maybe indicate in plot?]}}
\label{fig:specialJ1331}
\end{figure*}
